A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-10 calculus.tex
Results in paper

\begin{figure}[ht]
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[\EquationReasoning{}: $i = 1$]{
  i = 1
  \land n = 2
  \land \TransitionVar_2 = l_A - 3
  \land \TransitionVar_3 = 0
  \land \TransitionVar_5 = 1
  \land \TransitionVar_1 = \TransitionVar_4 = 1
  }
  \infer1[]{
    \begin{aligned}
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2
    \land l_A - i - 2 > 0 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume: \cref{sec:intuition:subsume}]{
    \begin{aligned}
    \Connected(A', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2 
    \land l_A - i - 2 > 0
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(A', \Filter) \land
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 
    \land n > i > 0 \\
    \land l_A - i - n > 0 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(A', \Filter)  
    \land i = \TransitionVar_5 
    \land n = \TransitionVar_4 + \TransitionVar_1 
    \land n > i > 0
    \land l_A - i - n > 0
    \land \TransitionVar_4 = \TransitionVar_1 \\
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1
    \land 1 = \TransitionVar_3 + \TransitionVar_1 
    \land i > 0
    \land \TransitionVar_4 = \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand $\Sigma, \FlowEq$]{
    \begin{aligned}
    i = \TransitionVar_5 
    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
    \land n > i > 0
    \land l_A - i - n > 0
    \land n = \TransitionVar_4 + \TransitionVar_1\\
    \land \Connected(A', \Filter) 
    \land \FlowEq(A', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \FlowEq(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{A'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{
    \Image{}_{A', \Map}(\Filter, \left[l_A, i, n\right]) 
    \land n \geq i 
    \land l_A - i - n > 0
    \land i > 0
    }
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $A'$ of \cref{fig:aut_a}. Note the liberal use of removal of equal
variables (e.g. if $x = y$ appears as a statement, we substitute $y$ for $x$
everywhere and drop $\ldots \land x=y \land \ldots$) to keep the size of the
equations managable.}\label{fig:derivation:single}
\end{figure}

>>>>

\begin{figure}[ht]
  \centering
\begin{prooftree}
  \hypo[]{{\text{Regular solving finishes the proof}}}
  \infer1[\EquationReasoning{}: $i = 1$]{
  i = 1
  \land n = 2
  \land \TransitionVar_2 = l_A - 3
  \land \TransitionVar_3 = 0
  \land \TransitionVar_5 = 1
  \land \TransitionVar_1 = \TransitionVar_4 = 1
  }
  \infer1[]{
    \begin{aligned}
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2
    \land l_A - i - 2 > 0 
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0 \\
    \end{aligned}}
  \infer1[\Subsume: \cref{sec:intuition:subsume}]{
    \begin{aligned}
    \Connected(A', \Filter) \land 
    \TransitionVar_1 = 1 \land 
    i = \TransitionVar_5 \land 
    n = 2 \land \\
    0 < i < 2 
    \land l_A - i - 2 > 0
    \land \TransitionVar_2 = l_A - i - 2 
    \land \TransitionVar_3 = 0
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $\TransitionVar_1 = 1$ \cref{sec:a_1}]{
    \begin{aligned}
    \Connected(A', \Filter) \land
    i = \TransitionVar_5 \land 
    n = 2\TransitionVar_1 
    \land n > i > 0 \\
    \land l_A - i - n > 0 
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: $=$-elimination]{
    \begin{aligned}
    \Connected(A', \Filter)  
    \land i = \TransitionVar_5 
    \land n = \TransitionVar_4 + \TransitionVar_1 
    \land n > i > 0
    \land l_A - i - n > 0
    \land \TransitionVar_4 = \TransitionVar_1 \\
    \land \TransitionVar_2 = l_A - i - n 
    \land \TransitionVar_3 = 1 - \TransitionVar_1
    \land 1 = \TransitionVar_3 + \TransitionVar_1 
    \land i > 0
    \land \TransitionVar_4 = \TransitionVar_1 
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand $\Sigma, \FlowEq$]{
    \begin{aligned}
    i = \TransitionVar_5 
    \land l_A = \sum_{k = 1}^{k = 5} \TransitionVar_k 
    \land n > i > 0
    \land l_A - i - n > 0
    \land n = \TransitionVar_4 + \TransitionVar_1\\
    \land \Connected(A', \Filter) 
    \land \FlowEq(A', \Filter)
    \end{aligned}
  }
  \infer1[\EquationReasoning{}: expand sum]{
    \Connected(A', \Filter) \land
    \FlowEq(A', \Filter) \land
    \left[l_A, i, n\right] =
    \sum\limits_{\Transition \in \Transitions_{A'} } \Map(\Transition) \cdot \Filter(\Transition)
  }
  \infer1[\Expand{}]{
    \Image{}_{A', \Map}(\Filter, \left[l_A, i, n\right]) 
    \land n \geq i 
    \land l_A - i - n > 0
    \land i > 0
    }
\end{prooftree}
\caption{A derivation tree for an application of
\cref{const:s1-in-c-dd,const:s1-substring,const:something-before-and-after,const:more-inside-than-before}
from \cref{ex:string-constraints}, corresponding to handling the Parikh
automaton $A'$ of \cref{fig:aut_a}. Note the liberal use of removal of equal
variables (e.g. if $x = y$ appears as a statement, we substitute $y$ for $x$
everywhere and drop $\ldots \land x=y \land \ldots$) to keep the size of the
equations manageable.}\label{fig:derivation:single}
\end{figure}

### 2023-07-10 evaluation.tex
Results in paper

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than lazy.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{0.75\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
\end{figure}

\subsubsection{Scalability}\label{sec:scaling}%
To get an idea of how the different backends scale, we additionally execute
\numprint{2000} different randomly sampled benchmarks with a
\numprint{120}-second timeout. In this set we also include two additional
variants of \Calculus{}: one with clause learning and restarts disabled, and one
with only clause learning disabled to show the impact of
\cref{sec:random-restarts}, and \cref{sec:clause-learning} respectively.

A cactus plot showing the number of instances solved within a given timeout can
be seen in \cref{fig:cactus}. We see that \Calculus{} outperforms \Nuxmv{} in
general, but that \Nuxmv{} might have better scaling properties on longer runs.
For machines with better single-core performance than our experiment machine
\Nuxmv{}'s scaling benefits kicks in faster, so it is likely that the amount of
CPU cycles spent on the problem is the underlying factor. This is likely due to
two factors. First, \Nuxmv{} is more mature than \Calculus{}, and it makes sense
that its more general model checking methods would pay off on more difficult
instances where more problem-specific methods offer less payoff. Second, for
longer running calculations clause learning as described in
\cref{sec:clause-learning} can be assumed to matter more. However, the clauses
we learn are still relatively unpolished, and in particular generalising
insights across product computations is nontrivial and therefore partially left
as future work.

>>>>

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than lazy.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \numprint{120}s-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
\end{figure}

\subsubsection{Scalability}\label{sec:scaling}%
To get an idea of how the different backends scale, we additionally execute
\numprint{2000} randomly sampled (without replacement) benchmarks with a
\numprint{120}-second timeout. In addition to the previous back-ends, we also
execute \Calculus{} with clause learning and restarts disabled, and with only
clause learning disabled to show the impact of \cref{sec:random-restarts}, and
\cref{sec:clause-learning} respectively.

A cactus plot showing the number of instances solved within a given timeout can
be seen in \cref{fig:cactus}. We see that \Calculus{} outperforms \Nuxmv{} in
general but that \Nuxmv{} might scale better on very long runtimes.
Non-systematic observations suggest that \Nuxmv{}'s scaling benefits kicks in
faster on newer machines, so it is likely that the amount of CPU cycles spent is
the underlying factor. This is likely due to two factors. First, \Nuxmv{} is
more mature than \Calculus{}, and it makes sense that its more general model
checking methods would pay off on more difficult instances where more
problem-specific methods offer less payoff. Second, for longer running
calculations clause learning as described in \cref{sec:clause-learning} can be
assumed to matter more. However, the clauses we learn are still unpolished, and
in particular, generalise poorly across product computations. Improvements are
left as future work.

### 2023-07-10 implementation.tex
Results in paper

We baseline using the same Presburger solver (\Princess{}), input file parser,
and automaton implementation as \Catra. We do this to better analyse the impact
of the calculus rules themselves. Using the formula of
\eqref{eq:generate-parikh}, we produce quantified Presburger formulae for each
successive term and add them to \Princess. We compute the product incrementally
term by term, checking satisfiability at each step. We use a priority queue to
select automata for each step and order them by their number of transitions. We
use this heuristic to put off computing large (and therefore slow) products until
we have to, hoping to find an empty intermittent product.

>>>>

We baseline using the same Presburger solver (\Princess{}), input file parser,
and automaton implementation as \Catra. We do this to better analyse the impact
of the calculus rules themselves. Adapting~\cite{generate-parikh-image}, we
produce quantified Presburger formulae for each successive term and add them to
\Princess. We compute the product incrementally term by term, checking
satisfiability at each step. We use a priority queue to select automata for each
step and order them by their number of transitions. We use this heuristic to put
off computing large (and therefore slow) products until we have to, hoping to
find an empty intermittent product. This is roughly similar to the approach
taken in~\cite{approximate-parikh}.
